---
title: "How to use LLMs to improve specialized ML tools"
subtitle: "The case of criminal offense classification"
author: "Dylan Pieper"
format: 
  revealjs:
    theme: white
    slide-number: true
    transition: fade
    margin: 0.05
    width: 1200
    height: 900
    css: custom.css
---

```{r setup, echo = FALSE}
green <- "darkgreen"
orange <- "#DE6F00"
red <- "#CD5C5C"
```

## The dilemma 😫
Are specialized ML tools and LLMs in competition, or can they work together?

:::: {.columns}

::: {.column width="50%"}
**Specialized tools**:

::: {.fragment fragment-index=1}
- Better for specific tasks
:::
::: {.fragment fragment-index=2}
- Validated, deterministic
:::
::: {.fragment fragment-index=3}
- Straightforward to retrain
:::
:::

::: {.column width="50%"}
**LLMs**:

::: {.fragment fragment-index=1}
- Flexible for many tasks
:::
::: {.fragment fragment-index=2}
- Not validated, undeterministic
:::
::: {.fragment fragment-index=3}
- Difficult to retrain
:::
:::

::::

## My use case 🔎

::: {.incremental}
- Classify 1,537 offense descriptions for a racial disparities study in Pennsylvania
  - Get **UCCS crime type** (violent, public order, property, criminal traffic, drug, DUI; [Uniform Crime Classification Standard](https://www.science.org/doi/10.1126/sciadv.abq8123))
  - Get harm-based measures ([UK's Office for National Statistics](https://www.ons.gov.uk/peoplepopulationandcommunity/crimeandjustice/articles/theimpactofcrimeonvictimsandsociety/march2022))
    - Harm level (e.g., individual vs. institutional)
    - Harm type (e.g., physical vs. financial or economic)
:::

## {data-background-image="img/chatbot_black_mirror.png" data-background-size="contain"}

## Goals 🎯

::: {.incremental}
1. **Evaluation**: 
  - Use LLM to discover biases and limitations of specialized tool through model disagreement and uncertainty
  - Use specialized tool as baseline to evaluate LLM performance through distribution overlap (PA offenses)
  
2. **Extension**: Use LLMs to get data not otherwise available
:::

## Tools 🛠️

::: {.incremental}
- **TOC**: 
  - Text-based Offense Classification (specialized ML tool; [University of Michigan](https://cjars-toc.isr.umich.edu))
  - Matches text to 258 codes ("murder") to get type ("violent")
:::

::: {.incremental}
- **GPT-5**: LLM by [OpenAI](https://platform.openai.com/docs/models)
  - ellmer's [parallel_chat()](https://ellmer.tidyverse.org/reference/parallel_chat.html) function
  - ellmer's [structured data](https://ellmer.tidyverse.org/articles/structured-data.html) feature
:::

## TOC results

Data TOC was "good" at classifying:

```{r, echo = FALSE}
toc <- openxlsx::readWorkbook("cjars_toc.xlsx", sheet = 2)
toc_uccs <- openxlsx::readWorkbook("cjars_toc.xlsx", sheet = 3)

selected_examples <- c(
  "Incest of Minor - Complainant Under 13 Years",
  "Invol. Deviate Sexual Intercourse W/Child",
  "Murder Of The First Degree",
  "Retail Theft-Take Mdse"
)

toc_preview <- toc |>
  dplyr::mutate(dplyr::across(where(is.numeric), ~round(.x, 3))) |>
  dplyr::left_join(toc_uccs, by = "uccs_code") |>
  dplyr::mutate(
    description = stringr::str_squish(description),
    charge_desc = tolower(charge_desc),
    offense_type_desc = tolower(offense_type_desc)
  ) |>
  dplyr::select(description, charge_desc, offense_type = offense_type_desc, probability) |>
  dplyr::filter(description %in% selected_examples) |>
  dplyr::arrange(description)
  
DT::datatable(toc_preview, 
   colnames = c(
    "Description" = "description",
    "UCCS Code" = "charge_desc",
    "Off. Type" = "offense_type", 
    "Probability" = "probability"
  ),
  options = list(
    scrollX = TRUE, 
     pageLength = 10, 
     dom = 't',
     ordering = FALSE
   )
) |>
  DT::formatStyle(columns = 1:ncol(toc_preview), fontSize = '20px')
```

**Mean confidence: `r scales::percent(mean(readxl::read_excel("cjars_toc.xlsx", sheet = 2) |> dplyr::pull(probability)), accuracy = 1)`**

## TOC results (cont'd)

Data TOC was "bad" at classifying:

```{r, echo = FALSE}
toc <- openxlsx::readWorkbook("cjars_toc.xlsx", sheet = 2)
toc_uccs <- openxlsx::readWorkbook("cjars_toc.xlsx", sheet = 3)

selected_examples <- c(
  "Aggravated Cruelty to Animals - Torture",
  "IDSI Person W/Mental Disability",
  "Incest of Minor - Complainant 13-18 Years",
  "Unlawful To Fire Any Weapon/Firearm/rocket In City"
)

toc_preview <- toc |>
  dplyr::mutate(dplyr::across(where(is.numeric), ~round(.x, 3))) |>
  dplyr::left_join(toc_uccs, by = "uccs_code") |>
  dplyr::mutate(
    description = stringr::str_squish(description),
    charge_desc = tolower(charge_desc),
    offense_type_desc = tolower(offense_type_desc)
  ) |>
  dplyr::select(description, charge_desc, offense_type = offense_type_desc, probability) |>
  dplyr::filter(description %in% selected_examples) |>
  dplyr::arrange(description)
  
DT::datatable(toc_preview, 
   colnames = c(
    "Description" = "description",
    "UCCS Code" = "charge_desc",
    "Off. Type" = "offense_type", 
    "Probability" = "probability"
  ),
  options = list(
    scrollX = TRUE, 
     pageLength = 10, 
     dom = 't',
     ordering = FALSE
   )
) |>
  DT::formatStyle(columns = 1:ncol(toc_preview), fontSize = '20px')
```

::: {style="font-size: 0.4em;"}
IDSI = Involuntary Deviate Sexual Intercourse
:::

**Mean confidence: `r scales::percent(mean(readxl::read_excel("cjars_toc.xlsx", sheet = 2) |> dplyr::pull(probability)), accuracy = 1)`**

## LLM results

Data LLM was "good" at classifying:

```{r, echo = FALSE}
oai_dat <- utils::read.csv("schema_openai.csv") |>
  dplyr::select(-harm_score, -action_type) |>
  dplyr::relocate(description)

DT::datatable(oai_dat |>
  dplyr::filter(description %in% selected_examples) |>
  dplyr::arrange(description) |>
  dplyr::mutate(probability = 1 - uncertainty_score) |>
  dplyr::select(-uncertainty_score),
  colnames = c(
    "Description" = "description",
    "Off. Category" = "offense_category",
    "Off. Type" = "offense_type",
    "Harm Level" = "harm_level",
    "Harm Type" = "harm_type",
    "Probability" = "probability"
  ),
  options = list(
    scrollX = TRUE, 
    pageLength = 10, 
    dom = 't',
    ordering = FALSE
  )
) |>
  DT::formatStyle(columns = 1:ncol(oai_dat), fontSize = '20px')
```

**Mean confidence: `r scales::percent(mean(1 - oai_dat$uncertainty_score), accuracy = 1)`**

## Model agreement

```{r, echo = FALSE}
offense_types <- toc |>
  dplyr::left_join(toc_uccs, by = "uccs_code") |>
  dplyr::select(description, offense_type_desc, probability) |>
  dplyr::mutate(
    offense_type_desc = tolower(offense_type_desc),
    description = stringr::str_squish(description)
  ) |>
  dplyr::rename(toc_offense_type = offense_type_desc) |>
  dplyr::left_join(
    oai_dat |> dplyr::mutate(description = stringr::str_squish(description)),
    by = "description"
  ) |>
  dplyr::rename(openai_offense_type = offense_type) |>
  dplyr::select(description, toc_offense_type, openai_offense_type, probability, uncertainty_score) |>
  dplyr::mutate(
    both_confident = probability >= 0.8 & uncertainty_score <= 0.2,
    agreement = toc_offense_type == openai_offense_type,
    review = dplyr::case_when(
      both_confident & !agreement ~ "yes, disagree & certain",
      !both_confident & !agreement ~ "yes, disagree & uncertain",
      TRUE ~ "no"
    )
  ) |>
  dplyr::select(-both_confident, -agreement)

review_summary <- offense_types |>
  dplyr::count(review) |>
  dplyr::mutate(
    percent = n / sum(n),
    review_clean = dplyr::case_when(
      review == "no" ~ "Agree",
      review == "yes, disagree & uncertain" ~ "Disagree & uncertain",
      review == "yes, disagree & certain" ~ "Disagree & certain"
    ),
    review_clean = forcats::fct_reorder(review_clean, n, .desc = TRUE)
  )

ggplot2::ggplot(review_summary, ggplot2::aes(x = review_clean, y = percent, fill = review_clean)) +
  ggplot2::geom_col(width = 0.7) +
  ggplot2::geom_text(
    ggplot2::aes(label = paste0(scales::comma(n), "\n(", scales::percent(percent, accuracy = 0.1), ")")),
    vjust = -0.5,
    size = 6,
    fontface = "bold",
    color = "black"
  ) +
  ggplot2::scale_y_continuous(
    labels = scales::percent_format(),
    expand = ggplot2::expansion(mult = c(0, 0.35))
  ) +
  ggplot2::scale_fill_manual(
    values = c(
      "Agree" = green,
      "Disagree & uncertain" = orange,
      "Disagree & certain" = red
    )
  ) +
  ggplot2::labs(
    title = NULL,
    x = NULL,
    y = NULL
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "none",
    plot.title = ggplot2::element_text(size = 20, face = "bold", hjust = 0.5),
    axis.text.x = ggplot2::element_text(size = 16),
    axis.text.y = ggplot2::element_text(size = 14),
    panel.grid.major.x = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank()
  )
```

## Topics of disagreement 

```{r, echo = FALSE}
# summary disagreement table by offense type
disagreement_by_type <- offense_types |>
  dplyr::group_by(toc_offense_type) |>
  dplyr::summarise(
    total_cases = dplyr::n(),
    disagreements = sum(toc_offense_type != openai_offense_type),
    disagreement_rate = disagreements / total_cases,
    .groups = 'drop'
  ) |>
  dplyr::arrange(desc(disagreement_rate)) |>
  dplyr::mutate(
    disagreement_rate_pct = scales::percent(disagreement_rate, accuracy = 0.1)
  ) |>
  dplyr::select(-disagreement_rate)

DT::datatable(
  disagreement_by_type,
  colnames = c(
    "Off. Type" = "toc_offense_type",
    "Total Cases" = "total_cases", 
    "Disagreements" = "disagreements",
    "Disagreement Rate" = "disagreement_rate_pct"
  ),
  options = list(
    scrollX = TRUE, 
    pageLength = 10, 
    dom = 't',
    ordering = FALSE
  )
) |>
  DT::formatStyle(columns = 1:ncol(disagreement_by_type), fontSize = '25px')
```

## Public order disagreements

Use a comparison table to facilitate review:

```{r, echo = FALSE}
review_examples <- offense_types |>
  dplyr::filter(review != "no") |>
  dplyr::select(description, toc_offense_type, openai_offense_type, review) |>
  dplyr::mutate(
    review_type = dplyr::case_when(
      review == "yes, disagree & certain" ~ "Disagree, certain",
      review == "yes, disagree & uncertain" ~ "Disagree, uncertain"
    )
  ) |>
  dplyr::select(description, toc_offense_type, openai_offense_type, review_type) |>
  dplyr::filter(toc_offense_type %in% "public order") |>
  dplyr::sample_n(8)
  dplyr::arrange(description)

DT::datatable(
  review_examples,
  colnames = c("Description", "TOC", "LLM", "Review"),
  options = list(
    scrollX = TRUE, 
    pageLength = 10, 
    dom = 't',
    ordering = FALSE
  )
) |>
  DT::formatStyle(
    "review_type",
    backgroundColor = DT::styleEqual(
      c("Disagree, certain", "Disagree, uncertain"),
      c(red, orange)
    ),
    color = "white",
    fontWeight = "bold"
  ) |>
  DT::formatStyle(columns = 1:ncol(review_examples), fontSize = '20px')
```

## Bias detection
I also discovered systematic biases:

```{r, echo = FALSE}
idsi_toc_num <- offense_types |>
  dplyr::filter(stringr::str_detect(description, "IDSI")) |>
  dplyr::filter(toc_offense_type == "violent") |>
  nrow()

idsi_openai_num <- offense_types |>
  dplyr::filter(stringr::str_detect(description, "IDSI")) |>
  dplyr::filter(openai_offense_type == "violent") |>
  nrow()

idsi_denom <- offense_types |>
  dplyr::filter(stringr::str_detect(description, "IDSI")) |>
  dplyr::filter(openai_offense_type == "violent") |>
  nrow()

ani_cruel_toc_num <- offense_types |>
  dplyr::filter(stringr::str_detect(description, "Animal")) |>
  dplyr::filter(toc_offense_type == "violent") |>
  nrow()

ani_cruel_openai_num <- offense_types |>
  dplyr::filter(stringr::str_detect(description, "Animal")) |>
  dplyr::filter(openai_offense_type == "violent") |>
  nrow()

ani_cruel_denom <- offense_types |>
  dplyr::filter(stringr::str_detect(description, "Animal")) |>
  dplyr::filter(openai_offense_type == "violent") |>
  nrow()
```

::: {.fragment}
**Rape misclassified ("IDSI")**
:::

::: {.incremental}
- TOC classified `r idsi_toc_num` of `r idsi_denom` as violent
- LLM classified `r idsi_openai_num` of `r idsi_denom` as violent
:::

::: {.fragment}
**Animal cruelty misclassified ("Animal")**
:::

::: {.incremental}
- TOC classified `r ani_cruel_toc_num` of `r ani_cruel_denom` as violent
- LLM classified `r ani_cruel_openai_num` of `r ani_cruel_denom` as violent
:::

## {data-background-image="img/chatbot_dogs.png" data-background-size="contain"}
## {data-background-image="img/archie.jpg" data-background-size="contain"}

## IDSI comparison

```{r, echo = FALSE}
review_examples <- offense_types |>
  dplyr::select(description, toc_offense_type, openai_offense_type, review) |>
  dplyr::mutate(
    review_type = dplyr::case_when(
      review == "no" ~ "Agree",
      review == "yes, disagree & certain" ~ "Disagree, certain",
      review == "yes, disagree & uncertain" ~ "Disagree, uncertain"
    )
  ) |>
  dplyr::select(description, toc_offense_type, openai_offense_type, review_type) |>
  dplyr::filter(stringr::str_detect(description, "IDSI")) |>
  dplyr::arrange(description) |>
  dplyr::slice_head(n = 10)

DT::datatable(
  review_examples,
  colnames = c("Description", "TOC", "LLM", "Review"),
  options = list(
    scrollX = TRUE, 
    pageLength = 10, 
    dom = 't',
    ordering = FALSE
  )
) |>
  DT::formatStyle(
    "review_type",
    backgroundColor = DT::styleEqual(
      c("Agree", "Disagree, certain", "Disagree, uncertain"),
      c(green, red, orange)
    ),
    color = "white",
    fontWeight = "bold"
  ) |>
DT::formatStyle(columns = 1:ncol(review_examples), fontSize = '20px')
```

## Hybrid framework

**Not an binary decision**

::: {.incremental}
1. **Initial tool**: Develop specialized ML tool as a baseline

2. **Auditing signal**: Use LLMs to identify disagreements, uncertain cases, and systematic biases

3. **Human review**: Use model comparison to focus review and model retraining
:::

## Implications

**The bias detection capability alone justifies LLM costs for criminal justice applications**

::: {.fragment}
**For ML tool users:**
:::

::: {.incremental}
- Audit your classifications
- Flag cases needing expert review
- Report biases and limitations
  - In your own work
  - To tool developers
:::

## {data-background-image="img/ostrich_black_box.png" data-background-size="contain"}

## Implications (cont'd)

**For ML tool developers:**

::: {.incremental}
- Audit your tools using model disagreement methods
- Create bias elimination and retraining workflows
- Leave behind audit trails and write reports
- If you have money, post bias bounties
:::

## Thank you️ ☺️

**Questions?**

::: {.fragment}
**Contact**: Dylan Pieper

- dylanpieper@gmail.com 
- [dylanpieper.github.io](https://dylanpieper.github.io)

**Repo**: [github.com/dylanpieper/posit-conf-2025](https://github.com/dylanpieper/posit-conf-2025)
:::